\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-25 paper/manuscript.qmd
submit

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  We introduce the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments, offering a better fit for data from subjective rating scales for which the two extremes correspond to discrete choices (e.g., "positive" *vs.* "negative", "true" *vs.* false", "real" *vs.* "AI-generated", ...), and for which the distance from the center might reflect a gradual process, such as confidence or certainty in that choice. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (the right side of the scale over the left side) and the confidence associated with each choice. To demonstrate its usage, we apply this model to two datasets (N=141 and N=189) in which participants judged whether face images were real photographs or AI-generated images, finding that facial attractiveness increased the likelihood of being judged as "real", particularly for male participants. As AI-generated content becomes increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically and practically urgent. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring, zero-and-one inflated Beta regression"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

The first sample includes the same participants as study 1 (N = 141, see above). 
For the second sample (data and preprocessing information available at https://github.com/RealityBending/FakeFace2), 248 participants were initially recruited through academic platforms (SONA and SurveySwap). We removed 15 participants with data suggesting low-effort responding (no variation in their responses indicative of straightlining) as well as those (N=16) that did not believe in the experimental manipulation and were fully confident that all images were real or fake.
<!-- 8 (3.23%) participants with no variation in their responses and 7 (2.92%) participants with no coherence (r < 0.1) b -->
Similarly to the first sample, we included all hetero and bisexual participants and stimuli from the opposite gender, resulting in a final sample included 189 participants (Mean age = 28.4, SD = 14.0, range: [18, 69]; Sex: 76.2% females). The study was approved by the University of Sussex Ethics Board (ER/ST633/1).

@@

For the second sample, the procedure was relatively similar to that of the first sample (described above in Study 1) with a few key differences.

>>>>

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: "0000-0003-3306-4695"
    role:
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  We introduce the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments, offering a better fit for data from subjective rating scales for which the two extremes correspond to discrete choices (e.g., "positive" *vs.* "negative", "true" *vs.* false", "real" *vs.* "AI-generated"), and for which the distance from the center might reflect a gradual process, such as confidence or certainty in that choice. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (the right side of the scale over the left side) and the confidence associated with each choice. To demonstrate its usage, we apply this model to two datasets (N=141 and N=189) in which participants judged whether face images were real photographs or AI-generated images, finding that facial attractiveness increased the likelihood of being judged as "real", particularly for male participants. Given that AI-generated content is becoming increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically important and practically urgent. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring, zero-and-one inflated Beta regression"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

The first sample includes the same participants as study 1 (N = 141, see above). 
For the second sample (data and preprocessing information available at https://github.com/RealityBending/FakeFace2), 248 participants were initially recruited through academic platforms (SONA and SurveySwap). We removed 15 participants with data suggesting low-effort responding (no variation in their responses indicative of straightlining) as well as those (N=16) that did not believe in the experimental manipulation and were fully confident that all images were real or fake.
<!-- 8 (3.23%) participants with no variation in their responses and 7 (2.92%) participants with no coherence (r < 0.1) b -->
Similarly to the first sample, we included all hetero and bisexual participants and stimuli from the opposite gender, resulting in a final sample included 189 participants (Mean age = 28.4, SD = 14.0, range: [18, 69]; Sex: 76.2% females). The study was approved by the University of Sussex Ethics Board (ER/ST633/1), but not preregistered.

@@

The procedure for the first sample is described in study 1. For the second sample, the procedure was relatively similar to that of the first sample (described above in Study 1) with a few key differences.

### 2025-06-25 paper/manuscript.tex
submit

\abstract{We introduce the Choice-Confidence (CHOCO) model as a tool to
dissect the cognitive mechanisms underlying decision-making judgments,
offering a better fit for data from subjective rating scales for which
the two extremes correspond to discrete choices (e.g., ``positive''
\emph{vs.} ``negative'', ``true'' \emph{vs.} false'', ``real''
\emph{vs.} ``AI-generated'', \ldots), and for which the distance from
the center might reflect a gradual process, such as confidence or
certainty in that choice. Conceptualized as a mixture of two Beta
distributions, CHOCO simultaneously estimates the probability of
selecting one category (the right side of the scale over the left side)
and the confidence associated with each choice. To demonstrate its
usage, we apply this model to two datasets (N=141 and N=189) in which
participants judged whether face images were real photographs or
AI-generated images, finding that facial attractiveness increased the
likelihood of being judged as ``real'', particularly for male
participants. As AI-generated content becomes increasingly
indistinguishable from real stimuli, understanding how individuals form
beliefs about the reality of what they perceive is both theoretically
and practically urgent. We discuss how these findings relate to the
broader effect of facial attractiveness, and suggest future directions
for both the cognitive and psychometric application of the CHOCO
framework. }

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, leftrule=.75mm, left=2mm, rightrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, opacityback=0, arc=.35mm]

@@

The first sample includes the same participants as study 1 (N = 141, see
above). For the second sample (data and preprocessing information
available at https://github.com/RealityBending/FakeFace2), 248
participants were initially recruited through academic platforms (SONA
and SurveySwap). We removed 15 participants with data suggesting
low-effort responding (no variation in their responses indicative of
straightlining) as well as those (N=16) that did not believe in the
experimental manipulation and were fully confident that all images were
real or fake. Similarly to the first sample, we included all hetero and
bisexual participants and stimuli from the opposite gender, resulting in
a final sample included 189 participants (Mean age = 28.4, SD = 14.0,
range: {[}18, 69{]}; Sex: 76.2\% females). The study was approved by the
University of Sussex Ethics Board (ER/ST633/1).

>>>>

\abstract{We introduce the Choice-Confidence (CHOCO) model as a tool to
dissect the cognitive mechanisms underlying decision-making judgments,
offering a better fit for data from subjective rating scales for which
the two extremes correspond to discrete choices (e.g., ``positive''
\emph{vs.} ``negative'', ``true'' \emph{vs.} false'', ``real''
\emph{vs.} ``AI-generated''), and for which the distance from the center
might reflect a gradual process, such as confidence or certainty in that
choice. Conceptualized as a mixture of two Beta distributions, CHOCO
simultaneously estimates the probability of selecting one category (the
right side of the scale over the left side) and the confidence
associated with each choice. To demonstrate its usage, we apply this
model to two datasets (N=141 and N=189) in which participants judged
whether face images were real photographs or AI-generated images,
finding that facial attractiveness increased the likelihood of being
judged as ``real'', particularly for male participants. Given that
AI-generated content is becoming increasingly indistinguishable from
real stimuli, understanding how individuals form beliefs about the
reality of what they perceive is both theoretically important and
practically urgent. We discuss how these findings relate to the broader
effect of facial attractiveness, and suggest future directions for both
the cognitive and psychometric application of the CHOCO framework. }

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy
Field}{0000-0003-3306-4695}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, colback=white, breakable, arc=.35mm, bottomrule=.15mm, left=2mm, toprule=.15mm, opacityback=0]

@@

The first sample includes the same participants as study 1 (N = 141, see
above). For the second sample (data and preprocessing information
available at https://github.com/RealityBending/FakeFace2), 248
participants were initially recruited through academic platforms (SONA
and SurveySwap). We removed 15 participants with data suggesting
low-effort responding (no variation in their responses indicative of
straightlining) as well as those (N=16) that did not believe in the
experimental manipulation and were fully confident that all images were
real or fake. Similarly to the first sample, we included all hetero and
bisexual participants and stimuli from the opposite gender, resulting in
a final sample included 189 participants (Mean age = 28.4, SD = 14.0,
range: {[}18, 69{]}; Sex: 76.2\% females). The study was approved by the
University of Sussex Ethics Board (ER/ST633/1), but not preregistered.
